---
title: "Identifying differentially methylated genes"
author: "EL Strand"
output:
  github_document: default
  pdf_document:
    keep_tex: yes
  html_document:
    toc: yes
    toc_depth: 6
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

# Differentially Methylated Genes 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Input from this script: https://github.com/emmastrand/EmmaStrand_Notebook/blob/master/_posts/2021-10-21-KBay-Bleaching-Pairs-WGBS-Analysis-Pipeline.md#kbay-wgbs-methylation-analysis-pipeline

I'm following Hollie Putnam and Kevin Wong's pipelines. 
- https://github.com/kevinhwong1/Thermal_Transplant_Molecular/blob/main/scripts/WGBS_GM.Rmd

## Load libraries 

https://github.com/hputnam/Geoduck_Meth/blob/master/RAnalysis/Scripts/GM.Rmd

```{r, message=FALSE, warning=FALSE}

# BiocManager::install("goseq")

library(plyr)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(readxl)
library(plotrix) 
library(gridExtra)
library(seacarb) 
library(pheatmap)
library(tidyr)
library(gplots)
library(cowplot)
library(lsmeans)
library(data.table)
library(RColorBrewer)
library(randomcoloR)
#library(GSEABase)
library(ggpubr)
library(hrbrthemes)
library(viridis)
library(factoextra)
#library(ropls)
#library(mixOmics)
library(vegan)
library(Rmisc)
```

## Load DNA methylation files 

Below commands were only run use to produce data file and then not run again for computational purposes.

```{r}
# meth5x <- list.files(path = "data/WGBS/output/meth_counts_5x", 
#                      pattern = ".bed$", full.names=TRUE) %>% set_names(.) %>%
#   map_dfr(read.csv,.id="Sample.ID", header=FALSE, sep="\t", na.string="NA", stringsAsFactors = FALSE) %>%
#   dplyr::select(-c(V3,V7:V14)) 
# 
# colnames(meth5x) <- c("Sample.ID", "scaffold", "position","per.meth","meth","unmeth","gene")
# meth5x$gene <- gsub("ID=","",meth5x$gene) #remove extra characters
# meth5x$gene <- gsub("Parent=","",meth5x$gene) #remove extra characters
# meth5x$Sample.ID <- gsub("data/WGBS/output/meth_counts_5x/","",meth5x$Sample.ID) #remove extra characters
# meth5x$Sample.ID <- gsub("_5x_sorted.tab_gene_CpG_5x_enrichment.bed","",meth5x$Sample.ID) #remove extra characters 
# meth5x$Sample.ID <- as.character(meth5x$Sample.ID)
```

Load df created in code above 

```{r}
load("data/WGBS/meth5x.RData")
```

### Metadata files 

```{r}
meta <- read.csv("data/metadata/Molecular_metadata.csv") %>%
  mutate(Timepoint = if_else(Plug_ID == "2153", "12 hour", Timepoint)) %>%
  dplyr::rename(Sample.ID = Plug_ID) %>% 
  dplyr::select(Sample.ID, Species, Tank, Treatment, Temperature, CO2, Timepoint, Sample.Date) 
  
meta$Sample.ID <- as.character(meta$Sample.ID)

meta$Timepoint <- factor(meta$Timepoint, 
                           levels = c("0 hour", "6 hour", "12 hour", "30 hour",
                                      "1 week", "2 week", "4 week", "6 week",
                                      "8 week", "12 week", "16 week"))

pacuta_clade <- read.delim2("data/metadata/clade_annotations_Pacuta.txt", header = TRUE, sep="\t") %>%
  separate(., Sample, c("Species", "Treatment", "Timepoint", "Sample.ID")) %>% 
  dplyr::select(-Treatment, -Timepoint) %>% dplyr::select(Sample.ID, Clade)

ploidyinfo <- read.delim2("data/metadata/samples_Pacuta.annotations.txt", 
                          header=TRUE, sep="\t", na.strings=c("","NA")) %>% 
  dplyr::rename(., Sample.ID = plugid) %>% 
  mutate_if(is.integer, as.character) %>% 
  dplyr::select(Sample.ID, ploidy, group, treesplit) %>%
  #mutate(treesplit = coalesce(treesplit, "D")) %>% 
  #changing NA values to "D" in treesplit -- this is just so triploidy is split in 2 groups and diploidy is altogether 
  left_join(., pacuta_clade, by="Sample.ID")

meta <- dplyr::left_join(meta, ploidyinfo, by="Sample.ID")
```

### Merge dataframes together

```{r}
df5x <- full_join(meth5x, meta, by = "Sample.ID") %>%
  mutate(meth_exp_group = case_when(
    group == "Group1" ~ "T1",
    group == "Group2" ~ "T1",
    group == "Group3" ~ "T2",
    group == "Group4" ~ "T2",
    group == "Group5" ~ "D",
    group == "Group6" ~ "D",
    group == "Group7" ~ "ungroup",
    group == "Ungroup" ~ "ungroup")) %>% 
  subset(!group == "Group7" & !group == "Ungroup") %>%
  filter(!Sample.ID == "1225") %>% 
  filter(!Sample.ID == "2197") 

df5x %>% 
  dplyr::select(Sample.ID, meth_exp_group, per.meth) %>% 
  filter(!is.na(per.meth)) %>% dplyr::select(-per.meth) %>%
  filter(!Sample.ID == "1225") %>% 
  filter(!Sample.ID == "2197") %>%
  distinct() %>% write.csv("data/WGBS/output/meth_pattern_groups.csv")
```

## Binomial GLM to test for differentially methylated genes

### Filter for genes with >5 methylated positions

Come back to this value of a cut off of 5 methylated positions per gene.. Does this need to be higher? Test cut offs.

```{r}
meth_table5x <- df5x
meth_table5x$sample_gene <- paste0(meth_table5x$Sample.ID, meth_table5x$gene)
```

```{r}
meth_table5x_position_counts <- dplyr::count(meth_table5x, vars = c(sample_gene))
meth_table5x_position_counts <- meth_table5x_position_counts[which(meth_table5x_position_counts$n > 5), ]

meth_table5x_filtered <- meth_table5x[meth_table5x$sample_gene %in% meth_table5x_position_counts$vars,]
meth_table5x_filtered %>% write.csv("data/WGBS/output/allgenes5x_filtered.csv")
```

```{r}
meth_table5x_filtered <- read.csv("data/WGBS/output/allgenes5x_filtered.csv") %>% 
  dplyr::select(-X) %>% distinct()
```

### Run loop for GLM 

Create data frame to store results

```{r}
results5x <- data.frame()
results10x <- data.frame()
```

First subset the unique dataframes and second run the GLMs

```{r}
gs5x <- unique(meth_table5x_filtered$gene)
```

#### 5x loop

```{r}
for(i in 1:length(meth_table5x_filtered$gene)){
    
    #subset the dataframe gene by gene
    meth_table_filt1 <- subset(meth_table5x_filtered, gene ==gs5x[1])
    
    # fit glm position model
    fit2 <- glm(matrix(c(meth, unmeth), ncol=2) ~ meth_exp_group, 
               data=meth_table_filt1, family=binomial)
    a2 <- anova(fit2, test="LRT")

    # capture summary stats to data frame
    df2 <- data.frame(gene = meth_table_filt1[1,7], ###for treesplit [1,8] or at least double check this
                     pval.methgroup = a2$`Pr(>Chi)`[2],
                     stringsAsFactors = F)
    # bind rows of temporary data frame to the results data frame
    results5x <- rbind(results5x, df2)
}
```

Export df created. Run the read.csv() function below the export code to read back in those df. When running the script multiple times, you don't need to run the binomial function above. Skip that code and just read in results5x or 10x df. 

```{r}
results5x %>% write.csv("data/WGBS/output/results5x_methgroup.csv") #9,647 genes - this should match length of gs5x
#results5x %>% write.csv("data/WGBS/output/results5x_ploidy.csv") #9,647 genes - this should match length of gs5x
#results5x %>% write.csv("data/WGBS/output/results5x_treesplit.csv") #9,647 genes - this should match length of gs5x

results5x <- read.csv("data/WGBS/output/results5x_methgroup.csv", header = TRUE) %>% 
  dplyr::select(-X) %>% distinct() ## distinct() necessary in case the model above went through multiple times
```

Some genes return the error: `"glm.fit: fitted probabilities numerically 0 or 1 occurredglm.fit: fitted probabilities numerically 0 or 1". These will be filtered out of analysis later on` and `Error in `contrasts<-`(`*tmp*`, value = contr.funs[1 + isOF[nn]]) : contrasts can be applied only to factors with 2 or more levels`. 

## Calculating adjusted p-values for glm

```{r}
results5x[is.na(results5x)] <- 0

#results5x$adj.pval.treesplit <- p.adjust(results5x$pval.treesplit, method='BH')
#results5x$adj.pval.ploidy <- p.adjust(results5x$pval.ploidy, method='BH')
results5x$adj.pval.methgroup <- p.adjust(results5x$pval.methgroup, method='BH')
```

### subsetting new df to only those adjusted p value columns 

Treesplit: 3,852 genes significantly DMG out of 9,467 genes that had at least 5 methylated positions and appeared in all samples.

Ploidy: 1,399 genes significantly DMG out of 9,467 genes that had at least 5 methylated positions and appeared in all samples.

methgroup: X genes significantly DMG out of 9,467 genes that had at least 5 methylated positions and appeared in all samples.

```{r}
DMG_5x_sig <- results5x %>%
  dplyr::select(gene, adj.pval.methgroup) %>% #dplyr::select(gene, adj.pval.treesplit) %>% #adj.pval.ploidy
  mutate(across(2, round, 8)) %>%
  filter(adj.pval.methgroup < 0.050) #filter(adj.pval.treesplit < 0.050)

# DMG_5x_sig <- results5x %>% 
#   dplyr::select(gene, adj.pval.Temp, adj.pval.CO2, adj.pval.Time,
#                 adj.pval.Temp_x_CO2, adj.pval.Temp_x_Time, adj.pval.CO2_x_Time, 
#                 adj.pval.Temp_x_CO2_x_Time) %>%
#   mutate(across(2:8, round, 8)) %>% #five decimal places to get rid of scientific notation
#   filter(adj.pval.Temp < 0.050 | adj.pval.CO2 < 0.050 | adj.pval.Time < 0.050 | 
#          adj.pval.Temp_x_CO2 < 0.050 | adj.pval.Temp_x_Time < 0.050 | adj.pval.CO2_x_Time < 0.050 | 
#          adj.pval.Temp_x_CO2_x_Time < 0.050) # filter for the significant ones

#DMG_5x_sig %>% write.csv("data/WGBS/output/DMG_5x_sig_treesplit.csv")
#DMG_5x_sig %>% write.csv("data/WGBS/output/DMG_5x_sig_ploidy.csv")
DMG_5x_sig %>% write.csv("data/WGBS/output/DMG_5x_sig_methgroup.csv")
```



#### Dataframes at this point 

- `eggNOGG` = original gene annotation file 
- `OE` = output from CpG OE script with weakly and heavily methylated gene sets 
- `gene_meta` = merged metadata of genes 
- `df5x / df10x` = .bed files and metadata together
- `meth_table5x_filtered / meth_table10x_filtered` = output from filtering all genes with 5+ methylated positions; input for the glm model
- `gs5x / gs10x` = list of unique genes from filtered methylation table  
- `results5x / results10x` = output from glm model; contains gene names, pvalues and adjusted pvalues for differentially methylated  
- `DMG_5x_sig / DMG_5x_sig` = only adjusted pvalues from the results 5 and 10x dfs 

## Significantly DMG - weakly and heavily 

```{r}
DMG_5x_sig #3,852 for treesplit; 1,399 for ploidy ; 3,870 for methgroup
DMG_5x_sig <- left_join(DMG_5x_sig, OE, by="gene") 
filter(DMG_5x_sig, status == "weakly") ##3,374 genes for treesplit; 1,117 for ploidy ; 3,375 for methgroup
filter(DMG_5x_sig, status == "heavily") ##478 genes for treesplit; 282 for ploidy ; 495 for methgroup 
```

### Adding "loci status" to each loci as a category and filtering to only significant DMGs 

```{r}
meth_table5x_filtered <- meth_table5x_filtered %>% 
  mutate(loci_status = case_when(
          per.meth <= 10 ~ "unmethylated",
          per.meth > 10 & per.meth < 50 ~ "sparsely methylated",
          per.meth >= 50 ~ "methylated"))

## hist for per.meth distribution for DMGs
meth_table5x_filtered_sigDMG <- meth_table5x_filtered[meth_table5x_filtered$gene %in% DMG_5x_sig$gene,]
```

## Methylation level calculations  

### GENERAL HISTOGRAM OF CPG LOCI METHYLATION LEVEL

```{r}
### number of loci per sample = 103,629 unique loci for all; 44,366 for sig DMG
## sample size of D = 16; T1=20; T2=17 
single_sample <- meth_table5x_filtered2_sigDMG %>% select(position) %>% distinct()
count(unique(single_sample$Sample.ID))

#### GENERAL HISTOGRAM OF CPG LOCI METHYLATION LEVEL 
## hist for per.meth distribution for all genes
histall <- meth_table5x_filtered %>%
  ggplot(aes(x=per.meth, fill=meth_exp_group)) + theme_bw() +
  facet_grid(factor(loci_status, levels=c('unmethylated','sparsely methylated','methylated'))~meth_exp_group, scales="free") +
  geom_histogram(alpha=0.6, position = 'identity') +
  scale_fill_manual(values=c("skyblue3", "olivedrab4", "darkgreen")) +
  labs(fill="") +
  theme(strip.text.x = element_text(size = 12, color = "black", face = "bold"),
        strip.text.y = element_text(size = 12, color = "black", face = "bold")) +
  theme(strip.background = element_rect(color="black", fill="white", linewidth=0.5, linetype="solid"))

## hist for per.meth distribution for significant DMGs 
histDMG <- meth_table5x_filtered_sigDMG %>%
  ggplot(aes(x=per.meth, fill=meth_exp_group)) + theme_bw() +
  facet_grid(factor(loci_status, levels=c('unmethylated','sparsely methylated','methylated'))~meth_exp_group, scales="free") +
  geom_histogram(alpha=0.6, position = 'identity') +
  scale_fill_manual(values=c("skyblue3", "olivedrab4", "darkgreen")) +
  labs(fill="") +
  theme(strip.text.x = element_text(size = 12, color = "black", face = "bold"),
        strip.text.y = element_text(size = 12, color = "black", face = "bold")) +
  theme(strip.background = element_rect(color="black", fill="white", linewidth=0.5, linetype="solid"))

##### without unmethylated 
histun <- meth_table5x_filtered %>%
  filter(!loci_status == "unmethylated") %>%
  ggplot(aes(x=per.meth, fill=meth_exp_group)) + theme_bw() +
  facet_grid(~meth_exp_group, scales="free") +
  geom_histogram(alpha=0.6, position = 'identity') +
  scale_fill_manual(values=c("skyblue3", "olivedrab4", "darkgreen")) +
  labs(fill="") +
  theme(strip.text.x = element_text(size = 12, color = "black", face = "bold"),
        strip.text.y = element_text(size = 12, color = "black", face = "bold")) +
  theme(strip.background = element_rect(color="black", fill="white", linewidth=0.5, linetype="solid"))

# ggsave(filename="data/WGBS/output/figures/locipercent_histun.jpeg", 
#        plot=histun, dpi=300, width=5.5, height=2.5, units="in")
```

### Filtering out loci with zero's 

Filter out loci that have median < 10% across all samples. 

meth_table5x_filtered = 5,346,733 loci 
meth_table5x_filtered_sigDMG = 2,278,227 loci

meth_table5x_filtered2 = 229,661 loci in 1,976 genes (4,503 loci per sample)
meth_table5x_filtered2_sigDMG =  198,647 loci in 1,447 genes (3,895 loci per sample)

```{r}
meth_table5x_filtered2 <- meth_table5x_filtered %>% 
  group_by(position) %>%
  mutate(position.median = median(per.meth)) %>%
  filter(!position.median < 10) %>% ungroup()

meth_table5x_filtered_out <- meth_table5x_filtered %>% 
  group_by(position) %>%
  mutate(position.median = median(per.meth)) %>%
  filter(!position.median > 10) %>% ungroup()

#unique(meth_table5x_filtered2$gene)

## viewing one loci to see if the above function worked 
meth_table5x_filtered2 %>% filter(position == "3396")
meth_table5x_filtered %>% filter(position == "3396")

meth_table5x_filtered2_sigDMG <- meth_table5x_filtered2[meth_table5x_filtered2$gene %in% DMG_5x_sig$gene,]
meth_table5x_filteredout_sigDMG <- meth_table5x_filtered_out[meth_table5x_filtered_out$gene %in% DMG_5x_sig$gene,]

unique(meth_table5x_filtered2_sigDMG$gene)

meth_table5x_filtered2 %>% write.csv("data/WGBS/output/meth_table5x_filtered2.csv")
meth_table5x_filtered2_sigDMG %>% write.csv("data/WGBS/output/meth_table5x_filtered2_sigDMG.csv")

meth_table5x_filtered2 <- read.csv("data/WGBS/output/meth_table5x_filtered2.csv") %>% dplyr::select(-X) %>% distinct()
meth_table5x_filtered2_sigDMG <- read.csv("data/WGBS/output/meth_table5x_filtered2_sigDMG.csv") %>% dplyr::select(-X) %>% distinct()
```

Calculating median loci methylation % per gene per sample for all DMGs 

```{r}
## Median from the filtered2 above
gene.stats <- meth_table5x_filtered2 %>%
  dplyr::group_by(Sample.ID, gene) %>%
  mutate(median.gene = median(per.meth),
         mean.gene = mean(per.meth)) %>% ungroup() %>%
  dplyr::group_by(Sample.ID) %>%
  mutate(median.sample = median(median.gene),
         mean.sample = mean(mean.gene)) %>% ungroup()

#gene.stats %>% dplyr::select(gene) %>% distinct()

gene.stats.DMGs <- meth_table5x_filtered2_sigDMG %>%
  dplyr::group_by(Sample.ID, gene) %>%
  mutate(median.gene = median(per.meth, na.rm=TRUE),
         mean.gene = mean(per.meth)) %>% ungroup() %>%
  dplyr::group_by(Sample.ID) %>%
  mutate(median.sample = median(median.gene),
         mean.sample = mean(mean.gene)) %>% ungroup()

## 1,447 genes of the DMGs have median loci % methylation above 10 
gene.stats.DMGs %>% dplyr::select(gene) %>% distinct() #3,560 genes filtered out
gene.stats %>% dplyr::select(gene) %>% distinct() #9,292 genes filtered out
  
## plot 
gene.stats.plot <-  gene.stats.DMGs %>% 
    #select(Sample.ID, meth_exp_group, gene, median.gene, mean.gene) %>% distinct() %>%
    dplyr::select(Sample.ID, meth_exp_group, median.sample, mean.sample) %>% distinct() %>%
    ggplot(., aes(x=meth_exp_group, y=mean.sample, color=meth_exp_group)) + 
    geom_boxplot() +
    geom_jitter(alpha=0.5, width = 0.25) + 
    theme_bw() + ylab("Mean CpG loci % methylation per gene") +
    #facet_wrap(~gene_status, scales = "free_y") + 
    scale_colour_manual(values = c("skyblue3", "olivedrab4", "darkgreen")) +
    theme(strip.text.x = element_text(size = 12, color = "black", face = "bold"),
          strip.text.y = element_text(size = 12, color = "black", face = "bold")) +
    theme(strip.background = element_rect(color="black", fill="white", linewidth=0.5, linetype="solid"))

# ggsave(filename="data/WGBS/output/figures/permeth_out_mean_persample.jpeg", 
#        plot=gene.stats.plot, dpi=300, width=5, height=4, units="in")
```

Statistics on the above 

1225 
2197

```{r}
gene.stats.df <- gene.stats.DMGs %>% 
  dplyr::select(Sample.ID, meth_exp_group, median.sample, mean.sample) %>% distinct()
  
aov <- aov(mean.sample ~ meth_exp_group, data=gene.stats.df)
summary(aov)
TukeyHSD(aov)
```


# Plotting heatmap of DMGs with inidividual variability 

`ploidy.meth.means.sigDMG` = significant DMGs. Mean CpG loci % methylation per gene per sample. 

```{r}
## use for per sample heatmap 
heatmap.sample.df <- gene.stats %>% filter(!Sample.ID == "1225") %>% filter(!Sample.ID == "2197") %>%
    dplyr::select(Sample.ID, meth_exp_group, gene, mean.gene) %>% distinct() 

heatmap.sample.DMG.df <- gene.stats.DMGs %>% filter(!Sample.ID == "1225") %>% filter(!Sample.ID == "2197") %>%
    dplyr::select(Sample.ID, gene, meth_exp_group, mean.gene) %>% distinct() 

## use for ploidy group heatmap 
heatmap.group.df <- heatmap.sample.df %>% group_by(gene, meth_exp_group) %>%
  mutate(mean.group = mean(mean.gene)) %>% ungroup() %>% dplyr::select(-Sample.ID, -mean.gene) %>% distinct()

heatmap.group.DMG.df <- heatmap.sample.DMG.df %>% group_by(gene, meth_exp_group) %>%
  mutate(mean.group = mean(mean.gene)) %>% ungroup() %>% dplyr::select(-Sample.ID, -mean.gene) %>% distinct()

heatmap.sample.DMG.df %>% subset(meth_exp_group == "D") %>% dplyr::select(Sample.ID) %>% distinct()
```


```{r}
library("ComplexHeatmap")

All_data <- heatmap.sample.DMG.df %>% dplyr::select(-meth_exp_group) %>%
  #pivot_wider(names_from = c(meth_exp_group), values_from = mean.group)
  pivot_wider(names_from = c(Sample.ID), values_from = mean.gene)
All_mat <- as.matrix(All_data[,-1]) #create matrix and remove gene name 
group_order<-c("D", "T1", "T2")

pdf("data/WGBS/output/figures/Heatmap_samples_DMG.pdf")
#heatmap_df<-heatmap.2(All_mat, 
heatmap.2(All_mat, 
          cexCol = 1.5, 
          distfun = function(x) as.dist(1 - cor(t(x), use = "pa")), 
          hclustfun = function(x) hclust(x,method = 'average'),
          key.xtickfun = function() {
  breaks = pretty(parent.frame()$breaks)
  breaks = breaks[c(1,length(breaks))]
  list(at = parent.frame()$scale01(breaks),labels = breaks)},
  Colv=TRUE, 
  col= rev(RColorBrewer::brewer.pal(name = "RdYlBu", n = 10)), 
  density.info = "none", 
  #ColSideColors=col_labels,
  #colCol = col_labels,
  trace = "none", 
  scale = "row", 
  labRow = FALSE,
  labCol = FALSE,
  sepcolor="none",
  sepwidth=c(0.1,0.01),
  lmat = rbind(c(0,3),c(2,1),c(4,0)), keysize=1, key.par = list(cex=0.65), lhei=c(1.5,4,1), lwid = c(1.5,4))
dev.off()
```

```{r}
zscore_df <- as.data.frame(heatmap_df$carpet) ##carpet score is the zscore that goes into the heatmap
## but the above df is not ordered by dendrogram like the heatmap is

##sample order of heatmap 
colInd_df <- as.data.frame(heatmap_df$colInd)
sample_order <- All_data %>% select(-gene) %>%
  gather(., "Sample.ID", "value", 1:51) %>% select(-value) %>% distinct()
sample_order$`heatmap_df$colInd` <- c(1:51)
colInd_df <- left_join(colInd_df, sample_order, by="heatmap_df$colInd") 
sample_ploidyinfo <- gene.stats %>% select(Sample.ID, meth_exp_group) %>% distinct()
sample_ploidyinfo$Sample.ID <- as.character(sample_ploidyinfo$Sample.ID)
colInd_df <- left_join(colInd_df, sample_ploidyinfo, by = "Sample.ID")

# saving gene row ids to match to name and z score above 
rowInd_df <- as.data.frame(heatmap_df$rowInd)
rowInd_df$gene <- c(1:1447)
rowInd_df$gene <- paste0("V", rowInd_df$gene)

## testing z score output to double check order of the gene names above 
## number below is the first row under rowInd above
test <- All_data[1145,] 
test <- test %>% gather("Sample.ID", "value", 2:4) %>% na.omit() ##2:4 for meth_exp_group
test <- test %>% mutate(mean = mean(value),
                        stdev = sd(value),
                        zscore = (value-mean)/stdev)
## confirmed that V1 is the first row of rowInd_df which is row 642 in All_data  
## next steps is to bind rowInd_df with the row id and then the gene name from All_data
All_data_gene_names <- All_data 
All_data_gene_names$rowInd <- c(1:1447)
#All_data_gene_names$rowInd <- as.character(All_data_gene_names$rowInd)

zscore_modified <- zscore_df %>% 
  rownames_to_column(., var = "meth_exp_group") %>% na.omit() %>%
  gather("gene", "zscore", 2:1448) %>%
  left_join(., rowInd_df, by = "gene") %>%
  dplyr::rename(., rowInd = `heatmap_df$rowInd`) %>%
  left_join(., All_data_gene_names, by = ("rowInd")) %>% #select(-c(6:58)) %>%
  dplyr::rename(., gene_number = gene.x) %>%
  dplyr::rename(., gene = gene.y) %>% #left_join(., ploidyinfo, by=c("Sample.ID")) %>%
  # mutate(meth_exp_group = case_when(
  #   group == "Group1" ~ "T1",
  #   group == "Group2" ~ "T1",
  #   group == "Group3" ~ "T2",
  #   group == "Group4" ~ "T2",
  #   group == "Group5" ~ "D",
  #   group == "Group6" ~ "D"
  # )) %>%
  dplyr::rename(meth_D = D) %>%
  dplyr::rename(meth_T1 = T1) %>%
  dplyr::rename(meth_T2 = T2) %>%
  spread(meth_exp_group, zscore)
zscore_modified$rowInd <- as.character(zscore_modified$rowInd)
## now I can subset to the conditions I want 

# Genes that are hypermethylated in Triploidy 2 (and not in Triploidy 1 or Diploidy) = 122 genes
hypermeth_trip2 <- zscore_modified %>% filter(T2 > 1) 
# Genes that are hypermethylated in Triploidy 1 (and not in Triploidy 2 or Diploidy)  = 138 genes
hypermeth_trip1 <- zscore_modified %>% filter(T1 > 1) 
# Genes that are hypermethylated in Diploidy (and not in Triploidy 2 or Triploidy 1)  = 460 genes
hypermeth_dip <- zscore_modified %>% filter(D > 1) 

hypermeth_trip2 %>% write.csv("data/WGBS/output/DMG_hypermeth_triploidy2.csv")
hypermeth_trip1 %>% write.csv("data/WGBS/output/DMG_hypermeth_triploidy1.csv")
hypermeth_dip %>% write.csv("data/WGBS/output/DMG_hypermeth_diploidy.csv")

hypermeth_trip2 <- read.csv("data/WGBS/output/DMG_hypermeth_triploidy2.csv")
hypermeth_trip1 <- read.csv("data/WGBS/output/DMG_hypermeth_triploidy1.csv")
hypermeth_dip <- read.csv("data/WGBS/output/DMG_hypermeth_diploidy.csv")
```

Plotting the hypermeth levels above 

```{r}
#hypermeth_trip2_plot <- hypermeth_trip2 %>% 
#hypermeth_trip1_plot <- hypermeth_trip1 %>% 
hypermeth_dip_plot <- hypermeth_dip %>% 
  dplyr::select(gene, meth_D, meth_T1, meth_T2) %>%
  gather("meth_exp_group", "per.meth", 2:4) %>%
    # mutate(loci_status = case_when(
    #       per.meth <= 10 ~ "unmethylated",
    #       per.meth > 10 & per.meth < 50 ~ "sparsely methylated",
    #       per.meth >= 50 ~ "methylated")) %>%
  ggplot(., aes(x=meth_exp_group, y=per.meth, color=meth_exp_group)) + theme_bw() + xlab("") +
  geom_boxplot() + geom_jitter(size=1, alpha=0.3, width=0.2) + 
  #facet_wrap(~loci_status, scales = "free_y") +
  scale_colour_manual(values = c("skyblue3", "olivedrab4", "darkgreen")) +
  theme(strip.text.x = element_text(size = 10, color = "black", face = "bold"),
        strip.text.y = element_text(size = 10, color = "black", face = "bold")) +
  theme(strip.background = element_rect(color="black", fill="white", linewidth=0.5, linetype="solid"))

ggsave(filename="data/WGBS/output/figures/hypermeth_dip.jpeg", 
       plot=hypermeth_dip_plot, dpi=300, width=4, height=3, units="in")

## statistics on the above 
hypermeth_stats <- hypermeth_trip2 %>% 
  dplyr::select(gene, meth_D, meth_T1, meth_T2) %>%
  gather("meth_exp_group", "per.meth", 2:4)

aov <- aov(per.meth ~ meth_exp_group, data=hypermeth_stats)
summary(aov)
TukeyHSD(aov)
```


# Principal Components Analysis & PERMANOVA 

## ALL 

## Creating matrix table

```{r}
## Create summary means for all genes with at least 5 methylated positions
pca.means.df <- aggregate(per.meth ~ gene*Sample.ID, data=meth_table5x_filtered, FUN=mean)
pca.means.df2 <- pca.means.df[pca.means.df$gene %in% DMG_5x_sig$gene,]

# transform df so column names are gene names 
All_data2 <- pca.means.df %>% 
  pivot_wider(names_from = gene, values_from = per.meth) %>%
  na.omit()

rownames(All_data2) <- paste0(All_data2$Sample.ID)
```

## Running prcomp function 

```{r}
scaled_pca_all <-prcomp(All_data2[c(2:ncol(All_data2))], scale=FALSE, center=TRUE)
fviz_eig(scaled_pca_all)

coral_info <- All_data2[c(1)]

pca_data_all <- scaled_pca_all%>%
  broom::augment(coral_info)%>%
  group_by(Sample.ID)%>%
  mutate(PC1.mean = mean(.fittedPC1),
         PC2.mean = mean(.fittedPC2))
```

## Examine PERMANOVA results.

```{r}
# scale data
vegan_all <- scale(All_data2[c(2:ncol(All_data2))])

# PERMANOVA 
permanova_all <- adonis2(vegan_all ~ group*ploidy, 
                         data = pca_data_all, method='eu', permutations = 9999)
capture.output(permanova_all, file = "data/WGBS/output/statistics/ploidy_PERMANOVA_all.txt")
```

## Assemble plot with background points

```{r}
#adding percentages on axis
names(pca_data_all)[3] <- "PCA1"
names(pca_data_all)[4] <- "PCA2"
names(pca_data_all)[5] <- "PCA3"
names(pca_data_all)[6] <- "PCA4"
percentage_all <- round((scaled_pca_all$sdev^2) / sum((scaled_pca_all$sdev^2)) * 100, 2)
percentage_all <- paste(colnames(pca_data_all[3:55]), "(",paste(as.character(percentage_all), "%", ")", sep="") )

## adding treatment column to visualize 
pca_data_all$Sample.ID <- as.character(pca_data_all$Sample.ID)
pca_data_all <- left_join(pca_data_all, ploidyinfo, by = c("Sample.ID")) %>%
  left_join(., meta, by = c("Sample.ID", "ploidy", "group", "treesplit", "Clade")) %>%
  #subset(!group == "Group7" & !group == "Ungroup") %>%
  mutate(meth_exp_group = case_when(
    group == "Group1" ~ "T1",
    group == "Group2" ~ "T1",
    group == "Group3" ~ "T2",
    group == "Group4" ~ "T2",
    group == "Group5" ~ "D",
    group == "Group6" ~ "D",
    group == "Group7" ~ "ungroup",
    group == "Ungroup" ~ "ungroup"
  ))

PCA_plot <- ggplot(pca_data_all, aes(PCA1, PCA2, color=meth_exp_group)) + #, label=Sample.ID
  geom_point(size = 4, aes(shape = ploidy)) +
  scale_colour_manual(values = c("skyblue3", "olivedrab4", "darkgreen", "pink3")) + #;  "darkgreen"
  #scale_colour_manual(values = c("blue", "lightblue", "salmon", "red3")) +
  scale_shape_manual(values=c(21, 17)) +
  theme_classic()+
  #geom_label() +
  ylab(percentage_all[2])+
  xlab(percentage_all[1])+
  #labs(color = "Group", shape = "Ploidy") + 
  #labs(color = "Treatment", shape = "Timepoint") +
  theme(legend.text = element_text(size=18), 
        legend.position="right",
        plot.background = element_blank(),
        legend.title = element_text(size=20), 
        plot.margin = margin(1, 1, 1, 1, "cm"),
        axis.text = element_text(size=18), 
        title = element_text(size=25, face="bold"), 
        axis.title = element_text(size=18))

ggsave(filename="data/WGBS/output/figures/PCA_all_group_PerMeth.jpeg", plot=PCA_plot, dpi=300, width=9, height=7.5, units="in")
```

Plot meth percent for Figure 2. 

```{r}
## make sure above pca.means.df is run only for the entire dataset not the significant genes 
#pca.means.df.env <- aggregate(per.meth ~ Sample.ID, data=pca.means.df, FUN=mean)
pca.means.df.env <- summarySE(pca.means.df, measurevar = c("per.meth"), groupvars = c("Sample.ID"))
pca.means.df.env$Sample.ID <- as.character(pca.means.df.env$Sample.ID)
pca.means.df.env <- left_join(pca.means.df.env, meta, by = c("Sample.ID"))

pca.means.df.env.plot <- pca.means.df.env %>%
  filter(!is.na(ploidy)) %>%
  ggplot(., aes(x=Timepoint, y=per.meth, color=Treatment, shape=ploidy)) + #, label=Sample.ID
  geom_jitter(size=2.5, width=0.3) + scale_shape_manual(values=c(21, 17)) + #geom_label() +
  ylab("CpG % methylation") +
  # geom_errorbar(aes(ymin=per.meth-se, ymax=per.meth+se), 
  #               linewidth=0.5, width=.075, position=position_dodge(),
  #               linetype = "dotted") +
  theme_bw() + scale_colour_manual(values = c("blue", "lightblue", "salmon", "red3")) 
## 1225 pink top; 2197 blue top 

ggsave(filename="data/WGBS/output/figures/env_permeth_all.jpeg", plot=pca.means.df.env.plot, dpi=300, width=6, height=5, units="in")

pca.means.df.env2 <- left_join(pca.means.df, meta, by=c("Sample.ID"))
pca.means.df.env2 <- summarySE(pca.means.df.env2, measurevar = c("per.meth"), groupvars = c("Treatment", "Timepoint"))

pca.means.df.env.plot2 <- pca.means.df.env2 %>%
  filter(!is.na(Treatment)) %>%
  ggplot(., aes(x=Timepoint, y=per.meth, color=Treatment, group=Treatment)) + 
  geom_point(size=2.5) + geom_line() +
  ylab("CpG % methylation") +
  geom_errorbar(aes(ymin=per.meth-se, ymax=per.meth+se),linewidth=0.5, width=.075) +
  theme_bw() + scale_colour_manual(values = c("blue", "lightblue", "salmon", "red3")) 

ggsave(filename="data/WGBS/output/figures/env_permeth_all2.jpeg", plot=pca.means.df.env.plot2, dpi=300, width=6, height=5, units="in")

stats3 <- lm(per.meth ~ Timepoint*Temperature*CO2, 
               na.action=na.omit, data=pca.means.df.env)

qqPlot(residuals(stats3)) 
hist(residuals(stats3))

t1 <- aov(per.meth ~ Timepoint*Temperature*CO2, 
               na.action=na.omit, data=pca.means.df.env)

stats3.anova <- Anova(stats3, ddf="lme4", type='III')

library(multcomp)
emmeans(stats3, pairwise ~ Timepoint*Temperature*CO2)
```


