---
title: "Hyper and hypo- methylated genes"
author: "EL Strand"
output:
  github_document: default
  pdf_document:
    keep_tex: yes
  html_document:
    toc: yes
    toc_depth: 6
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

# Hyper and hypo-methylated genes based on z-score of heatmap 

## Load libraries 

```{r}
library(ggplot2)
library(tidyverse)
library(Biostrings)
library(strex)
library(rstatix)
library("ComplexHeatmap")
library(gplots)
library(RColorBrewer)
library(dynamicTreeCut)

set.seed(123)
```

## Load data

```{r}
load("data/WGBS/meth_table5x_filtered_sigDMG.RData")

meth_levels_gene <- meth_table5x_filtered_sigDMG %>%
  dplyr::group_by(gene) %>%
  reframe(median_meth = median(per.meth),
          mean_meth = mean(per.meth))

# Step 1: summarize by gene and group
gene_stats <- meth_table5x_filtered_sigDMG %>%
  dplyr::group_by(gene, Sample.ID, meth_exp_group) %>%
  summarise(
    median_meth = median(per.meth),
    mean_meth   = mean(per.meth),
    .groups = "drop"
  )

gene_stats2 <- gene_stats %>%
  dplyr::group_by(gene) %>%
  summarise(
    median_gene = median(median_meth),
    mean_gene = mean(mean_meth)
  )

genes_above10 <- gene_stats2 %>% filter(mean_gene > 10) %>% pull(gene) ## 904 genes
genes_below10 <- gene_stats2 %>% filter(mean_gene < 10) %>% pull(gene) ## 2,788 genes 

# test <- meth_table5x_filtered_sigDMG %>% subset(gene == "Pocillopora_acuta_HIv2___RNAseq.10002_t")

# Step 2: 
### identify genes where all groups have median > 0
# genes_to_remove <- gene_stats %>%
#   dplyr::group_by(gene) %>%
#   summarise(all_pos = all(median_meth == 0), .groups = "drop") %>%
#   filter(all_pos) %>%
#   pull(gene)

### PoA style, at least X number of corals in each group need to be >0 
# genes_to_remove <- gene_stats %>%
#   dplyr::group_by(gene) %>%
#   summarise(all_pos = all(median_meth == 0), .groups = "drop") %>%
#   filter(all_pos) %>%
#   pull(gene)

# Step 3: drop those genes
meth_levels_gene_above10 <- gene_stats %>% filter(gene %in% genes_above10)
meth_levels_gene_below10 <- gene_stats %>% filter(gene %in% genes_below10)

# length(unique(meth_levels$gene)) 
## 843 genes from removing genes with all three groups == 0% median meth

## Step 4: calculate by meth exp group 
gene_stats_bygroup <- gene_stats %>% dplyr::group_by(gene, meth_exp_group) %>%
  summarise(
    mean_group = mean(mean_meth)
  )

gene_stats_bygroup_above10 <- gene_stats_bygroup %>% filter(gene %in% genes_above10)
gene_stats_bygroup_below10 <- gene_stats_bygroup %>% filter(gene %in% genes_below10)
```

## Heatmaps by sample x gene 

Prepping heatmap

```{r}
par("mar")
par(mar=c(1,1,1,1))

col_labels_df <- meth_levels_gene_below10 %>% dplyr::select(Sample.ID, meth_exp_group) %>% distinct() %>%
  column_to_rownames(., var = "Sample.ID") %>%
  mutate(color = case_when(
    meth_exp_group == "D" ~ "skyblue3",
    meth_exp_group == "T1" ~ "olivedrab4",
    meth_exp_group == "T2" ~ "darkgreen"
  ))

col_labels <- col_labels_df$color

my_colors  <- c("#dec0f1", 
                "#ea698b", 
                "#b185db",
                "#944bbb",
                "#ff90b3",
                "#f4acb7")
```

#### Genes with mean below 10 

2,788 genes 

Row = gene
Column = sample

```{r}
## 1. Make wide matrix: genes × samples for the "below 10" genes
All_data_below10 <- meth_levels_gene_below10 %>%
  dplyr::select(-median_meth, -meth_exp_group) %>%
  tidyr::pivot_wider(names_from = Sample.ID, values_from = mean_meth)

All_mat_below10 <- as.matrix(All_data_below10[ , -1])   # numeric matrix
rownames(All_mat_below10) <- All_data_below10$gene      # add gene names

## 2. Cluster genes (rows) by correlation
dist_genes <- as.dist(1 - cor(t(All_mat_below10),
                              use = "pairwise.complete.obs"))
hc_genes   <- hclust(dist_genes, method = "average")

## 3. Optionally cluster samples (columns) by correlation
dist_samples <- as.dist(1 - cor(All_mat_below10,
                                use = "pairwise.complete.obs"))
hc_samples   <- hclust(dist_samples, method = "average")

## 5. Draw clustered heatmap
pdf("data/figures/Heatmap_samples_DMG_below_mean10.pdf")

heatmap_df<-
  heatmap.2(All_mat_below10,
          cexCol = 1.5, 
          distfun = function(x) as.dist(1 - cor(t(x), use = "pa")), 
          hclustfun = function(x) hclust(x,method = 'average'),
          key.xtickfun = function() {
  breaks = pretty(parent.frame()$breaks)
  breaks = breaks[c(1,length(breaks))]
  list(at = parent.frame()$scale01(breaks),labels = breaks)},
  Rowv = as.dendrogram(hc_genes),
  Colv = as.dendrogram(hc_samples),
  col= rev(RColorBrewer::brewer.pal(name = "RdYlBu", n = 10)), 
  density.info = "none", 
  ColSideColors=col_labels,
  dendrogram = "both",
  trace = "none", 
  scale = "row", 
  labRow = FALSE,
  labCol = FALSE,
  sepcolor="none",
  sepwidth=c(0.1,0.01),
  #lmat = rbind(4:3, 2:1),
    ### 5 = legend, 4 = col dend, 3 = row dend, 2 = heatmap, 1 = labels
  #lmat = rbind(c(4,3), c(0,1), c(2,1)),
  lmat = rbind(c(5,4), c(0,1), c(3,2)),
  keysize=0.1, 
  key.par = list(mgp = c(1.5, 0.5, 0), cex=0.7), 
  lhei=c(0.5,0.1,2),
  lwid = c(1.5,4))

dev.off()
```
#### Genes with mean above 10 

904 genes  

```{r}
## 1. Make wide matrix: genes × samples for the "above 10" genes
All_data_above10 <- meth_levels_gene_above10 %>%
  dplyr::select(-median_meth, -meth_exp_group) %>%
  tidyr::pivot_wider(names_from = Sample.ID, values_from = mean_meth)

All_mat_above10 <- as.matrix(All_data_above10[ , -1])   # numeric matrix
rownames(All_mat_above10) <- All_data_above10$gene      # add gene names

## 2. Cluster genes (rows) by correlation
dist_genes <- as.dist(1 - cor(t(All_mat_above10),
                              use = "pairwise.complete.obs"))
hc_genes   <- hclust(dist_genes, method = "average")

## 3. Optionally cluster samples (columns) by correlation
dist_samples <- as.dist(1 - cor(All_mat_above10,
                                use = "pairwise.complete.obs"))
hc_samples   <- hclust(dist_samples, method = "average")

## 5. Draw clustered heatmap
pdf("data/figures/Heatmap_samples_DMG_above_mean10.pdf")

heatmap_df<-
  heatmap.2(All_mat_above10,
          cexCol = 1.5, 
          distfun = function(x) as.dist(1 - cor(t(x), use = "pa")), 
          hclustfun = function(x) hclust(x,method = 'average'),
          key.xtickfun = function() {
  breaks = pretty(parent.frame()$breaks)
  breaks = breaks[c(1,length(breaks))]
  list(at = parent.frame()$scale01(breaks),labels = breaks)},
  Rowv = as.dendrogram(hc_genes),
  Colv = as.dendrogram(hc_samples),
  col= rev(RColorBrewer::brewer.pal(name = "RdYlBu", n = 10)), 
  density.info = "none", 
  ColSideColors=col_labels,
  dendrogram = "both",
  trace = "none", 
  scale = "row", 
  labRow = FALSE,
  labCol = FALSE,
  sepcolor="none",
  sepwidth=c(0.1,0.01),
  #lmat = rbind(4:3, 2:1),
    ### 5 = legend, 4 = col dend, 3 = row dend, 2 = heatmap, 1 = labels
  #lmat = rbind(c(4,3), c(0,1), c(2,1)),
  lmat = rbind(c(5,4), c(0,1), c(3,2)),
  keysize=0.1, 
  key.par = list(mgp = c(1.5, 0.5, 0), cex=0.7), 
  lhei=c(0.5,0.1,2),
  lwid = c(1.5,4))

dev.off()
```



## Heatmaps by group x gene 

#### Above 10 

Generate heatmap 

```{r}
# reshape data: genes × methylation groups
gene_stats_bygroup_above10_heatmap <- gene_stats_bygroup_above10 %>% ungroup() %>%
  pivot_wider(names_from = meth_exp_group, values_from = mean_group)

gene_stats_bygroup_above10_heatmap_mat <- as.matrix(gene_stats_bygroup_above10_heatmap %>% dplyr::select(-gene))
rownames(gene_stats_bygroup_above10_heatmap_mat) <- gene_stats_bygroup_above10_heatmap$gene

# cluster genes by correlation
dist_mat_above10 <- as.dist(1 - cor(t(gene_stats_bygroup_above10_heatmap_mat), use = "pairwise.complete.obs"))
hc_genes_above10 <- hclust(dist_mat_above10, method = "average")

# optional: cluster samples (columns) too
hc_samples_above10 <- hclust(as.dist(1 - cor(gene_stats_bygroup_above10_heatmap_mat, 
                                     use = "pairwise.complete.obs")), method = "average")

# assign gene clusters (k = number of groups desired)
k_above10 <- 6
clusters_above10 <- cutree(hc_genes_above10, k = k_above10)

# row_colors <- RColorBrewer::brewer.pal(k, "Set1")[clusters]
my_colors  <- c("#b185db", "#f4acb7", "#ea698b", "#dec0f1", "#ff90b3", "#944bbb")

row_colors <- my_colors[clusters_above10]

col_colors_map <- c(
  "D"  = "skyblue3",
  "T1" = "olivedrab4",
  "T2" = "darkgreen"
)

# Match colors to matrix column order
col_labels <- col_colors_map[colnames(gene_stats_bygroup_above10_heatmap_mat)]

# heatmap: cluster genes, optionally samples
pdf("data/figures/Heatmap_samples_DMG_above_mean10_group.pdf", width=3, height=9)
sub_heatmap_df_above10 <-
heatmap.2(
  gene_stats_bygroup_above10_heatmap_mat,
  Rowv = as.dendrogram(hc_genes_above10),
  Colv = as.dendrogram(hc_samples_above10),
  distfun = function(x) as.dist(1 - cor(t(x), use = "pairwise.complete.obs")),
  hclustfun = function(x) hclust(x, method = "average"),
  scale = "row",
  trace = "none",
  density.info = "none",
  col = rev(brewer.pal(name = "RdYlBu", n = 10)),
  RowSideColors = row_colors,
  ColSideColors = col_labels,
  labRow = FALSE,
  labCol = FALSE,
  key = TRUE,
  cexCol = 1.2,
  lwid <- c(2, 0.5),
  lhei = c(0.5, 3.5),
  dendrogram = "row"
)

dev.off()
```


Pull genes in the different clusters 

```{r}
## Column orders of sample IDs in heatmap  
sub_colInd_df <- as.data.frame(sub_heatmap_df_above10$colInd)
sub_sample_order <- gene_stats_bygroup_above10 %>% ungroup() %>% select(-gene, -mean_group) %>% distinct()
sub_sample_order$`sub_heatmap_df_above10$colInd` <- c(1:3)
sub_colInd_df <- left_join(sub_colInd_df, sub_sample_order, by="sub_heatmap_df_above10$colInd")

## Row orders of genes in heatmap 
# saving gene row ids to match to name and z score above 
sub_rowInd_df <- as.data.frame(sub_heatmap_df_above10$rowInd)
sub_rowInd_df$gene <- c(1:904)
sub_rowInd_df$gene <- paste0("V", sub_rowInd_df$gene)

sub_All_data_gene_names <- gene_stats_bygroup_above10 %>% 
  pivot_wider(names_from = meth_exp_group, values_from = mean_group)

sub_All_data_gene_names$rowInd <- c(1:904)

gene_clusters_above10 <- data.frame(gene = names(clusters_above10), cluster = clusters_above10) %>% 
  mutate(cluster = paste0("Cluster ", cluster))

meth_levels_clusters <- gene_stats_bygroup_above10 %>% left_join(., gene_clusters_above10) %>% 
  mutate(cluster = as.character(cluster))

meth_levels_clusters %>% dplyr::group_by(cluster) %>%
  reframe(n=n_distinct(gene))

meth_levels_clusters_ordered <- meth_levels_clusters %>%
  mutate(cluster = factor(cluster, levels = c("Cluster 4", "Cluster 1", "Cluster 5", 
                                              "Cluster 2", "Cluster 3", "Cluster 6")))

ggplot(meth_levels_clusters_ordered,
       aes(x = meth_exp_group, y = scale(mean_group), group = gene, color = cluster)) +
  stat_summary(aes(group = cluster), fun = mean, geom = "line", linewidth = 2) +
  theme_bw() +
  facet_wrap(~cluster, ncol = 1, scales = "free_y") +
  scale_color_manual(values = c("Cluster 4"= "olivedrab4", #"#dec0f1", 
                                "Cluster 1"= "#3A8213", # "#b185db",
                                "Cluster 6"= "darkgreen", # "#944bbb",
                                "Cluster 3"=  "grey50", # "#ea698b", 
                                "Cluster 5"= "skyblue3", # "#ff90b3", 
                                "Cluster 2"= "skyblue3" # "#f4acb7")
                     
                     )) +
  scale_y_continuous(expand = expansion(mult = c(0.25, 0.25))) +
  labs(x = "Ploidy Group",
       y = "Normalized Mean Methylation",
       color = "Cluster") +
  theme(
    strip.background = element_rect(fill = "white", color = "black", linewidth = 1),
    strip.text = element_text(size = 10, face = "bold", color = "black"),
    legend.position = "none",
    axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0), size=10, face="bold")
  )

#ggsave("data/figures/Directional_pattern_above10.png", width=12, height=2)
ggsave("data/figures/Directional_pattern_above10_v2.png", width=2, height=9)

ggplot(meth_levels_clusters_ordered,
       aes(x = meth_exp_group, y = scale(mean_group), group = gene, fill = meth_exp_group)) +
  stat_summary(aes(group = cluster), fun = mean, geom = "bar", linewidth = 2) +
  theme_bw() +
  facet_wrap(~cluster, ncol = 1, scales = "free_y") +
  scale_fill_manual(values = c("skyblue3", "olivedrab4", "darkgreen")) +
  scale_y_continuous(expand = expansion(mult = c(0.25, 0.25))) +
  labs(x = "Ploidy Group",
       y = "Normalized Mean Methylation",
       color = "Cluster") +
  theme(
    strip.background = element_rect(fill = "white", color = "black", linewidth = 1),
    strip.text = element_text(size = 10, face = "bold", color = "black"),
    axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0), size=10, face="bold")
  )

ggsave("data/figures/Directional_pattern_above10_barchart_scaled.png", width=3.5, height=10)
```

#### below 10 

Generate heatmap 

```{r}
# reshape data: genes × methylation groups
gene_stats_bygroup_below10_heatmap <- gene_stats_bygroup_below10 %>% ungroup() %>%
  pivot_wider(names_from = meth_exp_group, values_from = mean_group)

gene_stats_bygroup_below10_heatmap_mat <- as.matrix(gene_stats_bygroup_below10_heatmap %>% dplyr::select(-gene))
rownames(gene_stats_bygroup_below10_heatmap_mat) <- gene_stats_bygroup_below10_heatmap$gene

# cluster genes by correlation
dist_mat_below10 <- as.dist(1 - cor(t(gene_stats_bygroup_below10_heatmap_mat), use = "pairwise.complete.obs"))
hc_genes_below10 <- hclust(dist_mat_below10, method = "average")

# optional: cluster samples (columns) too
hc_samples_below10 <- hclust(as.dist(1 - cor(gene_stats_bygroup_below10_heatmap_mat, 
                                     use = "pairwise.complete.obs")), method = "average")

# assign gene clusters (k = number of groups desired)
k_below10 <- 6
clusters_below10 <- cutree(hc_genes_below10, k = k_below10)

# row_colors <- RColorBrewer::brewer.pal(k, "Set1")[clusters]
#my_colors_below  <- c("#944bbb", "#ea698b", "#f4acb7", "#ff90b3", "#dec0f1", "#b185db")
my_colors_below  <- c("darkgreen", "grey50", "skyblue3", "skyblue3", "olivedrab4", "#3A8213")

row_colors_below <- my_colors_below[clusters_below10]

col_colors_map <- c(
  "D"  = "skyblue3",
  "T1" = "olivedrab4",
  "T2" = "darkgreen"
)

# Match colors to matrix column order
col_labels_below <- col_colors_map[colnames(gene_stats_bygroup_below10_heatmap_mat)]

# heatmap: cluster genes, optionally samples
pdf("data/figures/Heatmap_samples_DMG_below_mean10_group.pdf", width=7, height=10)
sub_heatmap_df_below10 <-
heatmap.2(
  gene_stats_bygroup_below10_heatmap_mat,
  Rowv = as.dendrogram(hc_genes_below10),
  Colv = as.dendrogram(hc_samples_below10),
  distfun = function(x) as.dist(1 - cor(t(x), use = "pairwise.complete.obs")),
  hclustfun = function(x) hclust(x, method = "average"),
  scale = "row",
  trace = "none",
  density.info = "none",
  col = rev(brewer.pal(name = "RdYlBu", n = 10)),
  RowSideColors = row_colors_below,
  ColSideColors = col_labels_below,   # column color bars
  labRow = FALSE,
  labCol = FALSE,
  key = TRUE,
  cexCol = 1.2,
  dendrogram = "row"
)
dev.off()
```

Pull genes in the different clusters 

```{r}
## Column orders of sample IDs in heatmap  
sub_colInd_df <- as.data.frame(sub_heatmap_df_below10$colInd)
sub_sample_order <- gene_stats_bygroup_below10 %>% ungroup() %>% select(-gene, -mean_group) %>% distinct()
sub_sample_order$`sub_heatmap_df_below10$colInd` <- c(1:3)
sub_colInd_df <- left_join(sub_colInd_df, sub_sample_order, by="sub_heatmap_df_below10$colInd")

## Row orders of genes in heatmap 
# saving gene row ids to match to name and z score below 
sub_rowInd_df <- as.data.frame(sub_heatmap_df_below10$rowInd)
sub_rowInd_df$gene <- c(1:2788)
sub_rowInd_df$gene <- paste0("V", sub_rowInd_df$gene)

sub_All_data_gene_names <- gene_stats_bygroup_below10 %>% 
  pivot_wider(names_from = meth_exp_group, values_from = mean_group)

sub_All_data_gene_names$rowInd <- c(1:2788)

gene_clusters_below10 <- data.frame(gene = names(clusters_below10), cluster = clusters_below10) %>% 
  mutate(cluster = paste0("Cluster ", cluster))

meth_levels_clusters <- gene_stats_bygroup_below10 %>% left_join(., gene_clusters_below10) %>% 
  mutate(cluster = as.character(cluster))

meth_levels_clusters %>% dplyr::group_by(cluster) %>%
  reframe(n=n_distinct(gene))

meth_levels_clusters_ordered <- meth_levels_clusters %>%
  mutate(cluster = factor(cluster, levels = c("Cluster 5", "Cluster 6", "Cluster 1", 
                                              "Cluster 3", "Cluster 4", "Cluster 2")))

ggplot(meth_levels_clusters_ordered,
       aes(x = meth_exp_group, y = scale(mean_group), group = gene, color = cluster)) +
  stat_summary(aes(group = cluster), fun = mean, geom = "line", linewidth = 2) +
  theme_bw() +
  facet_wrap(~cluster, ncol = 1, scales="free_y") +
  scale_color_manual(values = c("Cluster 5"=  "olivedrab4", # "#dec0f1", 
                                "Cluster 6"=  "#3A8213" ,  # "#b185db",
                                "Cluster 1"=  "darkgreen",  # "#944bbb",
                                "Cluster 2"=  "grey50", # "#ea698b", 
                                "Cluster 4"= "skyblue3",  # "#ff90b3", 
                                "Cluster 3"=   "skyblue3" #"#f4acb7")
                     
                     )) +
  scale_y_continuous(expand = expansion(mult = c(0.25, 0.25))) +
  labs(x = "Ploidy Group",
       y = "Normalized Mean Methylation",
       color = "Cluster") +
  theme(
    strip.background = element_rect(fill = "white", color = "black", linewidth = 1),
    strip.text = element_text(size = 8, face = "bold", color = "black")
  )

ggsave("data/figures/Directional_pattern_below10.png", width=3, height=6)
```

############# PATTERNS FOR STRING 

Load protein file 

```{r}
### protein fasta file 
fasta_file <- "data/Pocillopora_acuta_HIv2.genes.pep.faa.gz"
protein_sequences <- readAAStringSet(fasta_file)
names(protein_sequences) <- sub("-.*", "", names(protein_sequences))
```

Above 10% methylation

```{r}
gene_clusters_above10 %>% write.csv("data/WGBS/GOenrich/Above10/Gene_cluster_list.csv")

Cluster1_genes_above10 <- gene_clusters_above10 %>% subset(cluster=="Cluster 1") %>% pull(gene) ##116 genes
Cluster2_genes_above10 <- gene_clusters_above10 %>% subset(cluster=="Cluster 2") %>% pull(gene) ##312 genes
Cluster3_genes_above10 <- gene_clusters_above10 %>% subset(cluster=="Cluster 3") %>% pull(gene) ##99 genes
Cluster4_genes_above10 <- gene_clusters_above10 %>% subset(cluster=="Cluster 4") %>% pull(gene) ##89 genes
Cluster5_genes_above10 <- gene_clusters_above10 %>% subset(cluster=="Cluster 5") %>% pull(gene) ##218 genes
Cluster6_genes_above10 <- gene_clusters_above10 %>% subset(cluster=="Cluster 6") %>% pull(gene) ##70 genes

Cluster1_above10_matched_sequences <- protein_sequences[names(protein_sequences) %in% Cluster1_genes_above10] 
Cluster2_above10_matched_sequences <- protein_sequences[names(protein_sequences) %in% Cluster2_genes_above10] 
Cluster3_above10_matched_sequences <- protein_sequences[names(protein_sequences) %in% Cluster3_genes_above10] 
Cluster4_above10_matched_sequences <- protein_sequences[names(protein_sequences) %in% Cluster4_genes_above10] 
Cluster5_above10_matched_sequences <- protein_sequences[names(protein_sequences) %in% Cluster5_genes_above10] 
Cluster6_above10_matched_sequences <- protein_sequences[names(protein_sequences) %in% Cluster6_genes_above10] 

writeXStringSet(Cluster1_above10_matched_sequences, "data/WGBS/GOenrich/Cluster1_above10_matched_sequences.fasta")
writeXStringSet(Cluster2_above10_matched_sequences, "data/WGBS/GOenrich/Cluster2_above10_matched_sequences.fasta")
writeXStringSet(Cluster3_above10_matched_sequences, "data/WGBS/GOenrich/Cluster3_above10_matched_sequences.fasta")
writeXStringSet(Cluster4_above10_matched_sequences, "data/WGBS/GOenrich/Cluster4_above10_matched_sequences.fasta")
writeXStringSet(Cluster5_above10_matched_sequences, "data/WGBS/GOenrich/Cluster5_above10_matched_sequences.fasta")
writeXStringSet(Cluster6_above10_matched_sequences, "data/WGBS/GOenrich/Cluster6_above10_matched_sequences.fasta")
```

Below 10% methylation

```{r}
gene_clusters_below10 %>% write.csv("data/WGBS/GOenrich/Below10/Gene_cluster_list.csv")

Cluster1_genes_below10 <- gene_clusters_below10 %>% subset(cluster=="Cluster 1") %>% pull(gene) ##621 genes
Cluster2_genes_below10 <- gene_clusters_below10 %>% subset(cluster=="Cluster 2") %>% pull(gene) ##477 genes
Cluster3_genes_below10 <- gene_clusters_below10 %>% subset(cluster=="Cluster 3") %>% pull(gene) ##779 genes
Cluster4_genes_below10 <- gene_clusters_below10 %>% subset(cluster=="Cluster 4") %>% pull(gene) ##240 genes
Cluster5_genes_below10 <- gene_clusters_below10 %>% subset(cluster=="Cluster 5") %>% pull(gene) ##309 genes
Cluster6_genes_below10 <- gene_clusters_below10 %>% subset(cluster=="Cluster 6") %>% pull(gene) ##362 genes

Cluster1_below10_matched_sequences <- protein_sequences[names(protein_sequences) %in% Cluster1_genes_below10] 
Cluster2_below10_matched_sequences <- protein_sequences[names(protein_sequences) %in% Cluster2_genes_below10] 
Cluster3_below10_matched_sequences <- protein_sequences[names(protein_sequences) %in% Cluster3_genes_below10] 
Cluster4_below10_matched_sequences <- protein_sequences[names(protein_sequences) %in% Cluster4_genes_below10] 
Cluster5_below10_matched_sequences <- protein_sequences[names(protein_sequences) %in% Cluster5_genes_below10] 
Cluster6_below10_matched_sequences <- protein_sequences[names(protein_sequences) %in% Cluster6_genes_below10] 

writeXStringSet(Cluster1_below10_matched_sequences, "data/WGBS/GOenrich/Cluster1_below10_matched_sequences.fasta")
writeXStringSet(Cluster2_below10_matched_sequences, "data/WGBS/GOenrich/Cluster2_below10_matched_sequences.fasta")
writeXStringSet(Cluster3_below10_matched_sequences, "data/WGBS/GOenrich/Cluster3_below10_matched_sequences.fasta")
writeXStringSet(Cluster4_below10_matched_sequences, "data/WGBS/GOenrich/Cluster4_below10_matched_sequences.fasta")
writeXStringSet(Cluster5_below10_matched_sequences, "data/WGBS/GOenrich/Cluster5_below10_matched_sequences.fasta")
writeXStringSet(Cluster6_below10_matched_sequences, "data/WGBS/GOenrich/Cluster6_below10_matched_sequences.fasta")
```

## Load data 

```{r}
# load("data/WGBS/heatmap/zscore.RData") ## zscore_df
# load("data/WGBS/heatmap/Heatmap_output.RData") ## heatmap_df
# load("data/WGBS/heatmap/Heatmap_input.RData") ## All_data
# 
# groups <- read.csv("data/metadata/meth_pattern_groups.csv") %>% dplyr::select(-X)
# groups$Sample.ID <- as.character(groups$Sample.ID)
```

## Calculating z-score and hypermethylayed / hypomethylated groups 

Ordering samples in df

```{r}
# ## Column orders of sample IDs in heatmap  
# colInd_df <- as.data.frame(heatmap_df$colInd)
# 
# sample_order <- All_data %>% select(-gene) %>%
#   gather(., "Sample.ID", "value", 1:51) %>% select(-value) %>% distinct()
# sample_order$`heatmap_df$colInd` <- c(1:51)
# 
# colInd_df <- left_join(colInd_df, sample_order, by="heatmap_df$colInd") %>% 
#   left_join(., groups, by = "Sample.ID")
# 
# ## Row orders of genes in heatmap 
# # saving gene row ids to match to name and z score above 
# rowInd_df <- as.data.frame(heatmap_df$rowInd)
# rowInd_df$gene <- c(1:3692)
# rowInd_df$gene <- paste0("V", rowInd_df$gene)
# 
# ## testing z score output to double check order of the gene names above 
# ## number below is the first row under rowInd above
# test <- All_data[269,] 
# test <- test %>% gather("Sample.ID", "value", 2:52) %>% na.omit() ##2:4 for meth_exp_group
# test <- test %>% mutate(mean = mean(value),
#                         stdev = sd(value),
#                         zscore = (value-mean)/stdev)
# 
# ## confirmed that V1 is the first row of rowInd_df which is row 642 in All_data  
# ## next steps is to bind rowInd_df with the row id and then the gene name from All_data
# All_data_gene_names <- All_data 
# All_data_gene_names$rowInd <- c(1:3692)
# #All_data_gene_names$rowInd <- as.character(All_data_gene_names$rowInd)
```

Subsetting this list based on the filtering

```{r}
# All_data_gene_names_subsetted <- All_data_gene_names %>%
#   filter(gene %in% meth_levels$gene)
```



```{r}
# zscore_modified <- zscore_df %>% 
#   rownames_to_column(., var = "Sample.ID") %>% na.omit() %>%
#   gather("gene", "zscore", 2:last_col()) %>%
#   left_join(., rowInd_df, by = "gene") %>%
#   dplyr::rename(., rowInd = `heatmap_df$rowInd`) %>%
#   #left_join(., All_data_gene_names %>% dplyr::select(rowInd, gene), by = ("rowInd")) %>% 
#   right_join(All_data_gene_names_subsetted %>% dplyr::select(rowInd, gene), ., by = ("rowInd")) %>%
#   
#   dplyr::rename(., gene_number = gene.y) %>%
#   dplyr::rename(., gene = gene.x) %>% 
#   left_join(., groups, by=c("Sample.ID")) 
# 
# ### calculating median zscore per group
# zscore_median <- zscore_modified %>% group_by(meth_exp_group, gene) %>%
#   mutate(med_zscore = median(zscore)) %>% dplyr::select(gene, meth_exp_group, med_zscore) %>% distinct() %>%
#   spread(meth_exp_group, med_zscore)
# 
# # Genes that are hypermethylated in Triploidy 2 (and not in Triploidy 1 or Diploidy) = 101 genes
# hypermeth_trip2 <- zscore_median %>% filter(T2 > 0.75) 
# # Genes that are hypermethylated in Triploidy 1 (and not in Triploidy 2 or Diploidy)  = 70 genes
# hypermeth_trip1 <- zscore_median %>% filter(T1 > 0.75) 
# # Genes that are hypermethylated in Diploidy (and not in Triploidy 2 or Triploidy 1)  = 415 genes
# hypermeth_dip <- zscore_median %>% filter(D > 0.75) 
# 
# # Genes that are hypomethylated in Triploidy 2 (and not in Triploidy 1 or Diploidy) = 272 genes
# hypometh_trip2 <- zscore_median %>% filter(T2 < -0.75) 
# # Genes that are hypomethylated in Triploidy 1 (and not in Triploidy 2 or Diploidy)  = 140 genes
# hypometh_trip1 <- zscore_median %>% filter(T1 < -0.75) 
# # Genes that are hypomethylated in Diploidy (and not in Triploidy 2 or Triploidy 1)  = 143 genes
# hypometh_dip <- zscore_median %>% filter(D < -0.75) 
# 
# hypermeth_trip2 %>% write.csv("data/WGBS/hyper-hypo methylation/DMG_hypermeth_triploidy2.csv", row.names = FALSE)
# hypermeth_trip1 %>% write.csv("data/WGBS/hyper-hypo methylation/DMG_hypermeth_triploidy1.csv", row.names = FALSE)
# hypermeth_dip %>% write.csv("data/WGBS/hyper-hypo methylation/DMG_hypermeth_diploidy.csv", row.names = FALSE)
# 
# hypometh_trip2 %>% write.csv("data/WGBS/hyper-hypo methylation/DMG_hypometh_triploidy2.csv", row.names = FALSE)
# hypometh_trip1 %>% write.csv("data/WGBS/hyper-hypo methylation/DMG_hypometh_triploidy1.csv", row.names = FALSE)
# hypometh_dip %>% write.csv("data/WGBS/hyper-hypo methylation/DMG_hypometh_diploidy.csv", row.names = FALSE)
```

Plotting the hyper-meth/hypo-meth levels above 

```{r}
# subset_gene_list <- hypometh_dip
# 
# ## statistics on the above 
# ngenes = nrow(subset_gene_list)
# hypermeth_stats <- subset_gene_list %>% gather("group", "zscore", 2:last_col())
# 
# aov <- aov(zscore ~ group, data=hypermeth_stats)
# summary(aov)
# TukeyHSD(aov)
# 
# subset_gene_list %>%
#     
#   gather("group", "zscore", 2:last_col()) %>%
#   ggplot(., aes(x=group, y=zscore, color=group)) + theme_bw() + 
#   labs(
#     x="Ploidy Group",
#     color="Ploidy Group"
#   ) +
#   ggtitle(
#     label = paste0("n = ", ngenes, " genes"),
#     subtitle = "ANOVA p-value = <2e-16"
#   ) +
#   geom_boxplot(outlier.shape=NA) + geom_jitter(size=1, alpha=0.3, width=0.2) +
#   
#   scale_colour_manual(values = c("skyblue3", "olivedrab4", "darkgreen")) +
#   
#   theme(legend.position = "none",
#         panel.background=element_rect(fill='white', colour='black'),
#         axis.text.y = element_text(size=10, color="grey30"),
#         axis.text.x = element_text(size=10, color="grey30"),
#         plot.title = element_text(size=10, color="grey55", face = "italic"),
#         plot.subtitle = element_text(size=10, color="grey55", face = "italic"),
#         axis.title.y = element_text(margin = margin(t = 0, r = 10, b = 0, l = 0), size=12, face="bold"),
#         axis.title.x = element_text(margin = margin(t = 10, r = 0, b = 0, l = 0), size=12, face="bold"))
# 
# ggsave(filename="data/figures/Supplemental Figure 4 Zscore_hypometh_diploidy.jpeg", dpi=300, width=3.5, height=5, units="in")
```



















