---
title: "Classifications for epigenes"
author: "HM Putnam"
date: "2026-06-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries

```{r}
library(biomaRt)
library(dplyr)
library(stringr)
library(readr)
library(readxl)
library(writexl)
library(tidyverse)
library(car)
library(Rmisc)
library(data.table)
library(ggh4x)
#BiocManager::install(c("BiocFileCache", "biomaRt"), update = TRUE, ask = FALSE)

set.seed(123)
```

## Load Fellous data

```{r}
emma_notes <- read_xlsx("data/gene expression/proteins_to_annotate.xlsx", sheet=2)
```


All of the P.acuta blast top hits = 473 protien names

ENSID from the fasta = 512 unique EN IDs and protein names

```{r}
epigenes_blast <- read.delim2("data/gene expression/Pocillopora_acuta.tblastn_top_hits.tsv",
                              header=F, col.names = c("protein", "gene", "pident", "length", "mismatch",
                                                      "gapopen", "qstart", "qend", "sstart", "send", 
                                                      "evalue", "bitscore", "sframe"))

length(unique(epigenes_blast$protein))

ENSID <- read.csv("data/gene expression/fasta_headers.csv") %>% 
  separate(header, into = c("protein", "note", "EN ID", "note2"), sep = " ") %>%
  dplyr::select(protein, `EN ID`)

length(unique(ENSID$`EN ID`)) ## 512
length(unique(ENSID$protein)) ## 512

fellous_data <- epigenes_blast %>% dplyr::select(gene, protein) %>%
  left_join(., ENSID) %>%
  mutate(method = "Fellous_et_al")

length(unique(fellous_data$gene))
```

### Create descriptions and enzyme names based on EN IDs in Fellous 

```{r}
ids <- fellous_data$`EN ID`
ids_unique <- unique(ids)

human_ids <- ids_unique[str_detect(ids_unique, "^ENSP")]
mouse_ids <- ids_unique[str_detect(ids_unique, "^ENSMUSP")]
zfish_ids <- ids_unique[str_detect(ids_unique, "^ENSDARP")]

# Make archived marts once
# human_mart <- useEnsembl(
#   biomart = "genes",
#   dataset = "hsapiens_gene_ensembl",
#   version = 113
# )

## I ran into trouble connecting to the server once I updated my R
## mirror is the host server with the same data; this was trial and error with different mirrors
human_mart <- useEnsembl(biomart="genes", dataset="hsapiens_gene_ensembl")
mouse_mart <- useEnsembl(biomart="genes", dataset="mmusculus_gene_ensembl", mirror="asia")
zfish_mart <- useEnsembl(biomart = "genes", dataset = "drerio_gene_ensembl", mirror="useast")

# Function now uses the mart you give it
get_map <- function(ids, mart){
  getBM(
    attributes = c(
      "ensembl_peptide_id",
      "ensembl_gene_id",
      "external_gene_name",
      "description",
      "go_id",
      "name_1006"
    ),
    filters = "ensembl_peptide_id",
    values = ids,
    mart = mart
  )
}

human <- get_map(human_ids, human_mart)
mouse <- get_map(mouse_ids, mouse_mart)
zfish <- get_map(zfish_ids, zfish_mart)

mapped <- bind_rows(
  human %>% mutate(species = "human"),
  mouse %>% mutate(species = "mouse"),
  zfish %>% mutate(species = "zebrafish")
)

words <- unique(mapped$name_1006)
ENID_annotation <- bind_rows(human, mouse, zfish) %>% distinct(ensembl_peptide_id, .keep_all = TRUE)

length(unique(ENID_annotation$ensembl_peptide_id))
### 469 EN IDs out of 473 were annotated in this process 

fellous_data <- fellous_data %>% 
  left_join(., ENID_annotation %>% 
              dplyr::select(`EN ID`=ensembl_peptide_id, external_gene_name, description, name_1006)
            )

length(unique(fellous_data$gene)) ## 327 genes for
length(unique(fellous_data$`EN ID`)) ## 473 proteins 

fellous_data %>% write_xlsx("data/gene expression/fellous_data_mappedEnsembl.xlsx")
## if ensembl connection fails again, read back in the xlsx above
```

## Load P.acuta gff annotation data

```{r}
gff <- read.delim2("data/gene expression/Pocillopora_acuta_HIv2.genes.EggNog_results.txt") %>%
  dplyr::select(gene=1, description=Description) %>%
  mutate(method = "gff")

length(unique(gff$gene)) # 16,785 genes 
```

## Classifying genes based on keywords 

Deciding on epigenetic keywords 

```{r}
### CLASSIFIED AS MODERATE
epi_keywords <- regex(
  paste(c(
    "epigen", "chromatin", "histone", "nucleosome",
    "DNA methyl", "methyltransferase", "demethylase",
    "acetyltransferase", "deacetylase",
    "Polycomb", "Trithorax", "SWI/SNF", "remodel",
    "heterochromatin", "euchromatin", "reader",
    "bromodomain", "chromodomain", "SET domain",
    "TET", "DNMT", "HDAC", "KMT", "KDM", "EZH",
    "PRC1", "PRC2", "MBD", "MeCP"
  ), collapse = "|"),
  ignore_case = TRUE
)


### CLASSIFIED AS STRONG "chromatin remodeling", "SWI", "SNF", "nucleosome"
epi_strong <- regex(
  paste(c(
    "DNA methyl", "histone", "methyltransferase", "demethylase", "acetyltransferase", 
    "deacetylase", "Polycomb", "Trithorax"
  ), collapse = "|"),
  ignore_case = TRUE
)
```

Classifying both dfs 

```{r}
fellous_data_classified <- fellous_data %>%
  mutate(
    epi_flag = str_detect(paste(description, name_1006), epi_keywords),
    evidence_strength = case_when(
      str_detect(paste(description, name_1006), epi_strong) ~ "Strong/direct",
      epi_flag ~ "Moderate",
      TRUE ~ "No clear evidence from GO/description"
    )) 

fellous_data_classified %>% group_by(evidence_strength) %>%
  reframe(count = n_distinct(gene))
```

```{r}
gff_classified <- gff %>%   
  mutate(
    epi_flag = str_detect(paste(description), epi_keywords),
    evidence_strength = case_when(
      str_detect(paste(description), epi_strong) ~ "Strong/direct",
      epi_flag ~ "Moderate",
      TRUE ~ "No clear evidence from GO/description"
    )) 

gff_classified %>% group_by(evidence_strength) %>%
  reframe(count = n_distinct(gene))
```


### Putting final list together

```{r}
both_df <- full_join(gff_classified, fellous_data_classified) %>%
  subset(!evidence_strength=="No clear evidence from GO/description") %>%
  arrange(gene)

both_df %>% group_by(evidence_strength) %>%
  reframe(count = n_distinct(gene))

strong_evidence_df <- both_df %>%
  subset(evidence_strength=="Strong/direct")

## 517 unique genes with moderate or strong evidence 
length(unique(both_df$gene))
```

## Running ANOVA 

```{r}
meta <- read.csv("data/metadata/meth_pattern_groups.csv") %>% dplyr::select(-X) %>%
  mutate(Sample.ID = as.character(Sample.ID)) 

Pacuta_data_filt <- read.csv("data/gene expression/GE_Pacuta_data_filt.csv") %>%
  filter(X %in% strong_evidence_df$gene) %>% dplyr::rename(gene=X) %>%
  gather("Sample.ID", "count", 2:last_col()) %>% 
  mutate(Sample.ID = gsub("X", "", Sample.ID)) %>%
  left_join(., meta, by = "Sample.ID") %>%
  group_by(Sample.ID) %>% filter(!is.na(count))

### genes we have data for
strong_evidence_df %>% subset(evidence_strength == "Strong/direct") %>%
  filter(gene %in% Pacuta_data_filt$gene) %>% dplyr::select(gene) %>% distinct()

length(unique(Pacuta_data_filt$gene))
```

### Step 1: Run ANOVA for all genes + BH FDR correction 

313 genes 

```{r}
### running ANOVA loop on all genes in this df to see which ones have significant expression differences
epigene_anova <- data.frame()

for (i in unique(Pacuta_data_filt$gene)){
  stats_subset <- Pacuta_data_filt %>% subset(gene==i)
  aov_model <- aov(count ~ meth_exp_group, data=stats_subset)
  ANOVA <- Anova(aov_model, type="III")
  
  pvalue <- as.data.table(ANOVA)[2,4] %>% dplyr::rename(pvalue=`Pr(>F)`)
  DF <- data.frame(gene=i, pvalue)
  
  epigene_anova <- rbind(epigene_anova, DF)
}

epigene_anova$padj <- p.adjust(epigene_anova$pvalue, method="BH")
```

### Step 2: Keep only significant genes

69 significant genes post padj

```{r}
sig_genes <- epigene_anova %>% filter(padj < 0.05)
Pacuta_data_filt_sig <- Pacuta_data_filt %>% filter(gene %in% sig_genes$gene)

length(unique(Pacuta_data_filt_sig$gene))
```

### Step 3: Run Tukey post hoc tests only on significant genes

```{r}
epigene_tukey <- data.frame()

for (i in sig_genes$gene){
  stats_subset <- Pacuta_data_filt %>% subset(gene==i)
  aov_model <- aov(count ~ meth_exp_group, data=stats_subset)
  TUKEY <- TukeyHSD(aov_model)
  
  Tukeydf <- as.data.frame(TUKEY$meth_exp_group[,4]) %>% 
    rownames_to_column("comparison") %>% 
    dplyr::rename(pvalue=`TUKEY$meth_exp_group[, 4]`)
  
  Tukeydf$gene <- i
  
  epigene_tukey <- rbind(epigene_tukey, Tukeydf)
}

epigene_tukey$padj <- p.adjust(epigene_tukey$pvalue, method="BH") 

epigene_results <- epigene_tukey %>% dplyr::select(-pvalue) %>%
  spread(comparison, padj) %>%
  mutate_if(is.numeric, round, 8)
```

```{r}
# epigene_results_filtered <- epigene_results %>% distinct() %>%
#   mutate(sig = ifelse(pvalue < 0.050, "significant", "NS"))
# 
# ## genes with significantly different expression = 140 genes 
# significant_genes = epigene_results_filtered %>% filter(pvalue < 0.050) %>% dplyr::select(gene) %>% distinct()

#### mean expression of significant genes 
# Pacuta_data_filt_significant <- Pacuta_data_filt 

pacuta_counts <- read.csv("data/gene expression/pacuta_counts_vst.csv") %>% dplyr::rename(gene=X) %>% 
  filter(gene %in% epigene_results$gene) %>% rename_with(~ sub("^X", "", .x)) %>%
  gather("Sample.ID", "count", 2:last_col()) %>%
  left_join(., meta)

length(unique(pacuta_counts$gene))

mean_expression <- summarySE(pacuta_counts, measurevar = c("count"),
                               groupvars = c("gene", "meth_exp_group")) %>%
  dplyr::select(-N, -sd, -se, -ci) %>%
  spread(meth_exp_group, count)
```

Adding strong evidence df annotations to the significant genes 

```{r}
epigene_results %>% left_join(., strong_evidence_df, by = "gene") %>% 
  write_xlsx("data/gene expression/Epigenes_classified_significant_9-30-2026_tobin.xlsx")

# sig_df_ <- both_df %>% 
#   filter(gene %in% significant_genes$gene) %>% left_join(., epigene_results_filtered, by = "gene")
# 
# ## 109 with strong/direct evidence 
# both_df_sig_strong <- both_df_sig %>% subset(evidence_strength == "Strong/direct")
# length(unique(both_df_sig_strong$gene))
```

## plotting strong only

Calculating distance from diploid 

```{r}
sig_PH <- epigene_results %>% 
  dplyr::select(gene, T1=`T1-D`, T2=`T2-D`) %>%
  gather("ploidy", "TPH", T1:T2) %>%
  filter(TPH < 0.050) 

diploid_distance <- mean_expression %>% #filter(gene %in% both_df_sig_strong$gene) %>%
  dplyr::select(gene, D, T1, T2) %>%
  mutate(
    T1 = T1-D,
    T2 = T2-D
  ) %>%
  gather("ploidy", "distance", T1:T2) %>% dplyr::select(-D)

gene_exp_stats_for_plot <- full_join(sig_PH, diploid_distance) %>%
  filter(!is.na(TPH))

## 67 genes to plot 
length(unique(gene_exp_stats_for_plot$gene))
```

Adding bin information

```{r}
# both_df_sig_strong %>% write_xlsx("data/gene expression/Epigenes_classified_sig_strong_7-9-2026_tobin.xlsx")
# 

# both_df_sig_strong_bin <- both_df_sig_strong %>%
#   mutate(
#     search_text = paste(description, name_1006, sep = " "),
#     bin = case_when(
#       str_detect(search_text, regex("tRNA", ignore_case = TRUE)) ~ "tRNA methyltransferase",
#       str_detect(search_text, regex("rRNA", ignore_case = TRUE)) ~ "rRNA methyltransferase",
#       str_detect(search_text, regex("lysine.*methyltransferase", ignore_case = TRUE)) ~ "Histone methyltransferase",
#       str_detect(search_text, regex("histone.*methyltransferase", ignore_case = TRUE)) ~ "Histone methyltransferase",
#       str_detect(search_text, regex("histone.*deacetylase", ignore_case = TRUE)) ~ "Histone deacetylase",
#       str_detect(search_text, regex("demethylase", ignore_case = TRUE)) ~ "Histone demethylase",
#       str_detect(search_text, regex("WD40 repeats", ignore_case = TRUE)) ~ "WD40 repeats",
#       str_detect(search_text, regex("PHD", ignore_case = TRUE)) ~ "PHD finger protein",
# 
#       # Epitranscriptome
#       str_detect(search_text, regex("mRNA", ignore_case = TRUE)) ~ "mRNA cap 0 methyltransferase",
#       str_detect(search_text, regex("RsmB NOP family", ignore_case = TRUE)) ~ "RsmB NOP family",
#       str_detect(search_text, regex("gTMT family", ignore_case = TRUE)) ~ "gTMT family",
# 
#       str_detect(search_text, regex("Homocysteine", ignore_case = TRUE)) ~ "Homocysteine S-methyltransferase",
# 
#       str_detect(search_text, regex("protein-arginine", ignore_case = TRUE)) ~
#         "Protein arginine N-methyltransferase activity",
#       str_detect(search_text, regex("Protein arginine", ignore_case = TRUE)) ~
#         "Protein arginine N-methyltransferase activity",
# 
#       str_detect(search_text, regex("DNA.*methyltransferase", ignore_case = TRUE)) ~ "DNA methyltransferase",
#       str_detect(search_text, regex("reader", ignore_case = TRUE)) ~ "DNA methylation reader",
#       str_detect(search_text, regex("Arsenite", ignore_case = TRUE)) ~ "Arsenite methyltransferase",
#       str_detect(search_text, regex("carnosine", ignore_case = TRUE)) ~ "Carnosine methyltransferase",
#       str_detect(search_text, regex("Thiopurine", ignore_case = TRUE)) ~ "Thiopurine S-methyltransferase",
#       str_detect(search_text, regex("carboxyl", ignore_case = TRUE)) ~ "Carboxyl methyltransferase",
#       str_detect(search_text, regex("o-methyltransferase", ignore_case = TRUE)) ~ "O-methyltransferase",
#       str_detect(search_text, regex("aminomethyltransferase", ignore_case = TRUE)) ~ "Aminomethyltransferase activity",
#       str_detect(search_text, regex("^(?!.*histone).*deacetylase", ignore_case = TRUE)) ~ "Deacetylase",
# 
#       TRUE ~ "methyltransferase"
#     )
#   ) %>%
#   select(-search_text)

both_df_sig_strong_bin <- read_xlsx("data/gene expression/Epigenes_classified_significant_9-30-2026_binned.xlsx")

both_df_sig_strong_bin %>% filter(!is.na(bin)) %>% dplyr::select(gene) %>% distinct()
```


Plotting

```{r}
final_df_for_plot <- gene_exp_stats_for_plot %>% 
  left_join(., both_df_sig_strong_bin %>% dplyr::select(gene, method, description, name_1006, bin)) %>%
  filter(!is.na(bin))

length(unique(final_df_for_plot$gene))

final_df_for_plot %>% subset(ploidy=="T2") %>% filter(distance>0) 

final_df_for_plot %>%
  
  #ggplot(aes(x = sign(distance) * log10(abs(distance) + 1), y = gene)) +
  ggplot(aes(x=distance, y=gene)) + 
  geom_point(aes(fill = ploidy), size = 2, shape = 21, alpha = 0.65, color = "black") +
  facet_grid2(
    bin ~ .,
    scales = "free",
    space = "free",
    labeller = labeller(bin = label_wrap_gen(width = 5))
  ) +
  ylab("") +
  scale_fill_manual(values = c("olivedrab4", "darkgreen")) +
  geom_vline(xintercept = 0, color = "grey60", lty = "dotted") +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    legend.position = "none",
    strip.text.x = element_text(face = "bold", size = 8),
    strip.text.y = element_text(face = "bold", size = 8, angle = 0),
    strip.background.y = element_blank(),
    strip.clip = "off"
  )

ggsave("data/figures/GE_epigenetic_machinery_distance_20260930.png", height=12, width=3)
```








