---
title: "Manuscript_figures"
author: "Emma Strand"
date: "2024-03-03"
output: html_document
---

# Figures for manuscript 

## load libraries 

```{r}
library(ggplot2)
library(ggsignif)
library(ggpubr)
library(cowplot)
```


## Figure 2

A) All genes with >5 positions of data

This PCA will be in DMG-Pacuta script from meth5xfiltered df 

left off at why the ungroup is left out of PCA.... columns are 55 but should be 61? re-run DMG-pacuta from start.

```{r}
treatment_pca <- ggplot(pca_data_all, aes(PCA1, PCA2, color=Treatment)) + #, label=Sample.ID
  geom_point(size = 4, aes(shape = Timepoint)) +
  scale_colour_manual(values = c("blue", "lightblue", "salmon", "red3")) +
  theme_bw()+
  ylab(percentage_all[2])+
  xlab(percentage_all[1])+
  labs(color = "Treatment", shape = "Timepoint") +
  theme(legend.text = element_text(size=12), 
        legend.position="right",
        plot.background = element_blank(),
        legend.title = element_text(size=12), 
        plot.margin = margin(1, 1, 1, 1, "cm"),
        axis.text = element_text(size=12),
        axis.title.x = element_text(margin = margin(t=10,r=0,b=0,l=0), size=12, face="bold"),
        axis.title.y = element_text(margin = margin(t=0,r=10,b=0,l=0), size=12, face="bold"))
```


B) Ploidy Group PCA: All genes with >5 positions of data

This PCA will be in DMG-Pacuta script from meth5xfiltered df 

```{r}
ploidy_pca <- ggplot(pca_data_all, aes(PCA1, PCA2, color=meth_exp_group)) + #, label=Sample.ID
  geom_point(size = 4, aes(shape = ploidy)) +
  scale_colour_manual(values = c("skyblue3", "olivedrab4", "darkgreen", "pink3")) +
  theme_bw()+
  scale_shape_manual(values=c(21, 17)) +
  ylab(percentage_all[2])+
  xlab(percentage_all[1])+
  labs(color = "Group", shape = "Ploidy") +
  theme(legend.text = element_text(size=12), 
        legend.position="right",
        plot.background = element_blank(),
        legend.title = element_text(size=12), 
        plot.margin = margin(1, 1, 1, 1, "cm"),
        axis.text = element_text(size=12),
        axis.title.x = element_text(margin = margin(t=10,r=0,b=0,l=0), size=12, face="bold"),
        axis.title.y = element_text(margin = margin(t=0,r=5,b=0,l=0), size=12, face="bold"))
```


C) gene.stats df from DMG-Pacuta 

Genes that are significantly differentially methylated, have >5 positions of data, and median >10%.

```{r}
fig2C <- gene.stats.DMGs %>% 
  subset(!meth_exp_group == "ungroup") %>%
    dplyr::select(Sample.ID, meth_exp_group, median.sample, mean.sample) %>% distinct() %>%
    ggplot(., aes(x=meth_exp_group, y=mean.sample, color=meth_exp_group)) + 
    geom_boxplot(outlier.shape=NA) +
    geom_jitter(size=3, alpha=0.7, width = 0.25) + 
    theme_bw() + 
    geom_signif(comparisons = list(c("D", "T1"), c("T1", "T2"), c("D","T2")), 
                annotations = "***", 
                y_position = c(65,61,67),
                color="grey30", 
                size = 0.5) + ## size = linewidth()
    ylab("CpG methylation level (%)") +
    #xlab("Ploidy Group") +
    xlab("") +
    labs(color = "Group") +
    scale_colour_manual(values = c("skyblue3", "olivedrab4", "darkgreen")) +
    scale_y_continuous(expand=expansion(mult=c(0.15,0.17))) +
    #annotate(geom = "text", x=0.5, y=0.9, label = "p < 0.0001", size=8, color="grey45") +
    theme(axis.text.x = element_text(size = 12, color = "black"),
          axis.text.y = element_text(size = 12, color = "black"),
          strip.background = element_rect(color="black", fill="white", linewidth=0.5, linetype="solid"),
          legend.position = "right",
          legend.text = element_text(size=12),
          axis.title.x = element_text(margin = margin(t=10,r=0,b=0,l=0), size=12, face="bold"),
          axis.title.y = element_text(margin = margin(t=0,r=10,b=0,l=0), size=12, face="bold"))
```

Heatmap

```{r}
par("mar")
par(mar=c(1,1,1,1))

col_labels_df <- heatmap.sample.DMG.df %>% dplyr::select(Sample.ID, meth_exp_group) %>% distinct() %>%
  column_to_rownames(., var = "Sample.ID") %>%
  mutate(color = case_when(
    meth_exp_group == "D" ~ "skyblue3",
    meth_exp_group == "T1" ~ "olivedrab4",
    meth_exp_group == "T2" ~ "darkgreen"
  ))
col_labels <- col_labels_df$color

pdf("data/WGBS/output/figures/Heatmap_samples_DMG.pdf")
heatmap_df<-
  heatmap.2(All_mat,
          cexCol = 1.5, 
          distfun = function(x) as.dist(1 - cor(t(x), use = "pa")), 
          hclustfun = function(x) hclust(x,method = 'average'),
          key.xtickfun = function() {
  breaks = pretty(parent.frame()$breaks)
  breaks = breaks[c(1,length(breaks))]
  list(at = parent.frame()$scale01(breaks),labels = breaks)},
  Colv=TRUE, 
  col= rev(RColorBrewer::brewer.pal(name = "RdYlBu", n = 10)), 
  density.info = "none", 
  ColSideColors=col_labels,
  dendrogram = "column",
  trace = "none", 
  scale = "row", 
  labRow = FALSE,
  labCol = FALSE,
  sepcolor="none",
  sepwidth=c(0.1,0.01),
  #lmat = rbind(4:3, 2:1),
    ### 5 = legend, 4 = col dend, 3 = row dend, 2 = heatmap, 1 = labels
  #lmat = rbind(c(4,3), c(0,1), c(2,1)),
  lmat = rbind(c(5,4), c(0,1), c(3,2)),
  keysize=0.1, 
  key.par = list(mgp = c(1.5, 0.5, 0), cex=0.7), 
  lhei=c(0.5,0.1,2),
  lwid = c(1.5,4))
dev.off()
```


Putting altogether 








